---
title: "Model a count-type response using a Poisson regression based CLAMP model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clamp-poisson}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 4, fig.align = "center", fig.cap = "&nbsp;", 
  dpi = 120
)
```

```{r setup, message=FALSE}
# library(clamp)
devtools::load_all(".")
library(matrixStats)
```

In this vignette, we show how to apply the `clamp()` function on a Poisson
regression based Clamp model dedicated to handling count-type responses.

## Get genotype data and generate count-type responses

We simulate response using genotype data from an R package named
[bigsnpr](https://github.com/privefl/bigsnpr/).
The explanatory variables `X` in this example, 
saved in `./data/genotype/genotype-subset-2.rds`, 
is a subset of genotype data containing 2000 successive SNPs of 500 individuals. 
The data are extracted using the following code: 

```{r, eval = F}
# install.packages("bigsnpr")
library(bigsnpr)

`%&%` <- function(a,b) paste0(a,b)
data.dir <- "../data/genotype"
dir.create(data.dir, recursive=TRUE, showWarnings=FALSE)

.bed.file <- data.dir %&% "1000G_phase3_common_norel.bed"
if (!file.exists(.bed.file)) {
  bigsnpr::download_1000G(data.dir)
}

.bk.file <- data.dir %&% "1000G_phase3_common_norel.rds"
if (!file.exists(.bk.file)){
  BED <- snp_readBed(.bed.file)
}
dat <- snp_attach(.bk.file)$genotypes

nn <- 500
pp <- 2000

set.seed(123456789)
startpoint <- sample(1 : (ncol(dat)-pp), size = 1)
if (nn < 2490) {ii.idx <- sample(1 : 2490, size = nn)}
X <- dat[ii.idx, startpoint : (startpoint+pp-1)]

example.data.dir <- "./example-data/"
saveRDS(X, file = example.data.dir %&% "genotype-subset-2.rds")
```

Three effect variables are randomly selected, each having a non-zero
effect size of 1 (before scaling). 
Then, we generate the linear predictor `eta`. 
We scale the linear combination to avoid unexpected large or small values in `eta`.

```{r}
`%&%` <- function(a,b) paste0(a,b)
example.data.dir <- "./example-data/"

X <- readRDS(example.data.dir %&% "genotype-subset-2.rds")
nn <- nrow(X)  # 500
pp <- ncol(X)  # 2000

n_effect_vars <- 3

set.seed(12345)

## independent variables with non-zero effects
effect_idx <- sample(1:pp, size = n_effect_vars)
print(effect_idx)

## effect size
effect_size <- rep(1, times = n_effect_vars)

## linear predictor
eta <- scale(X[,effect_idx, drop=F] %*% as.matrix(effect_size))

## response
y <- rpois(nn, exp(eta))
```

The histogram of the synthetic response is:

```{r}
hist(y, breaks = 12)
```

## Fit a Poisson regression based clamp model

To specify fitting a **Poisson regression based clamp model**, 
use the argument `family = "poisson"`. Correspondingly, the
`clamp()` function call is:

```{r}
res_cl <- clamp(X, y, family = "poisson", estimate_residual_variance = F)
```

When fitting a GLM-based clamp model, we prefer introducing an intercept term. 
By default, the `clamp()` function sets `intercept=TRUE` 
such that an intercept is automatically fitted in the model. 

By default, we assume there are no more than 10 effect variables
and set `maxL = min(10, ncol(X))`. You may choose to specify another
value to `maxL`.

By setting `estimate_residual_variance=F`, we fix the `residual_variance=1`.
Thus, the `clamp` fitted result is yielded from the "original" iterative 
reweighted least squared (IRLS) algorithm: 

$$
\hat{b}_j := \frac{\sum_{i=1}^n w_{i} x_{ij} y_i}{\sum_{i=1}^n w_{i} x_{ij}^2},
$$

$$
s^2_j := \frac{1}{\sum_{i=1}^n w_i x_{ij}^2},
$$

where, $w_{i}$ is the weight of the $i$th individual computed from IRLS, 
and $\hat{b}_j$ and $s_j^2$ are the maximum likelihood estimate 
and its variance of $b_j$ conditioning on $b_j \neq 0$, i.e., 
the $j$-th variable is an effect variable. 

## Check the variable selection results

### Posterior inclusion probabilities

The `summarize_coefficients()` function offers 10 (by default)
variables with the highest posterior inclusion probabilities (PIPs). By
default, we output 95% credible intervals for the posterior means of each
variable.

```{r}
summarize_coefficients(res_cl)
```

True effect variables (`X51`, `X142`, and `X720`) are detected with high PIPs. 



### Credible sets (CS)

As with the SuSiE method, we output 95% credible sets by default.

```{r}
summary(res_cl)
```

Each credible set contain one variable.

### Graphical Display

The `clamp_plot()` function provides plots for PIP of each variable.

By default, since the intercept is just an auxiliary variable, we do not
include it in the plot. In this case, argument `intercept_index` needs
to be specified. (P.S. By default, we set `include_intercept = FALSE` to
remove the intercept term on the plot.)

```{r}
clamp_plot(res_cl, y = "PIP", effect_indices = effect_idx)
```

By specifying `effect_indices`, the true effect variables are
colored in red. The 95% credible sets are identified as circled in
different colors.

## What if we let `estimate_residual_variance=T`? 

```{r}
res_cl2 <- clamp(X, y, family = "poisson", estimate_residual_variance = T)
```

Show the outputs: 

```{r}
summarize_coefficients(res_cl2)
clamp_plot(res_cl2, y = "PIP", effect_indices = effect_idx)
```

```{r}
print(paste("estimated residual variance:", res_cl2$sigma2))
```

In this case, the estimated residual variance is relatively close to one. 
Besides, the fitted result `res_cl2` is comparable with `res_cl`; 
nevertheless, the posterior SDs of the nonzero coefficients is larger 
in `res_cl2` than in `res_cl`. 

However, if we review *clamp-logistic.Rmd* (a hyper-link?), 
it is not certain whether setting `estimate_residual_variance=T` is always
a good choice. 


## Refine a Poisson regression based clamp model via robust estimation

As we know, count data (in the response) usually contains extremely
large values. These unexpected values may affect the model fitting,
resulting in false discoveries or even failure to fit the model.

Hence, we suggest robust estimation when modelling a Poisson
regression based clamp model. Our `clamp` function provides several
robust estimation approaches. Based on our experience in simulation
experiments, we recommend the **Huber reweighting method**, which, in each
iteration, reweights each data point by Huber weight[@huber1964]
according to its residual updated from the previous iteration [(by
M-estimation)]{style="color: gray;"}.

The corresponding function call is:

```{r}
res_cl_hb <- clamp(X, y, family = "poisson", estimate_residual_variance = F,
                   robust_method = "huber")
```

### Check the refined variable selection results

The fitted results as well as graphical display are:

```{r}
summarize_coefficients(res_cl_hb)
```

```{r}
summary(res_cl_hb)
```


```{r}
clamp_plot(res_cl_hb, y = "PIP", effect_indices = effect_idx)
```

By applying the robust Huber reweighting method, 
the null variable `X54` is detected along with the true effect variable `X51`, 
but the PIP of the former is relatively lower than that of the latter. 
The 95% credible interval of the posterior mean of `X54` contains, 
indicating that `X54` might not be a significant effect variable; 
it pops-up probabily because it is highly correlated with `X51`.
Besides, as suggested above, 
`X51` is captured by two different credible sets, 
one of which contains only `X51`. 
This may indicate that `X51` is likely to be more important than `X54`. 


Other than performing Huber reweighting via M-estimation procedures, we
also provide an option of Huber reweighting via
S-estimation[@rousseeuw1984]. The corresponding function call is:

```{r}
res_cl_hb2 <- clamp(X, y, family = "poisson", estimate_residual_variance = F,
                     robust_method = "huber", 
                     robust_estimator = "S")
```

The fitted results as well as graphical display are:

```{r}
summarize_coefficients(res_cl_hb2)
```

```{r}
summary(res_cl_hb2)
```


```{r}
clamp_plot(res_cl_hb2, y = "PIP", effect_indices = effect_idx)
```

In simulation experiments, 
this selection result is comparable with the result applying the 
Huber M-estimator. 


## Suggestions for preprocessing count-type responses

Two cases in the count data response that may result in
difficulty fitting a Poisson regression based clamp model. One is
the presence of too many zeros, and the other is the existence of
some unexpectedly large values. The former can be addressed by fitting a
zero-inflated Poisson or a negative binomial regression model;
however, the clamp versions of these two genearlized linear models have not 
yet been developed in our `clamp` package. 
The latter may lead to unexpected failure of fitting a clamp model, 
as those extreme values may lead to overflow in certain fitting procedure. 

A simple approach to simultaneously address both problems is to preprocess the response: 
we scale the response without changing the shape of its original distribution. 
Specifically, the following transformation is operated  

```{r, eval = F}
y <- exp(log1p(y))
```

and then the new $y$ is plugged into the `clamp` function to fit a clamp model. 
This transformation on $y$ is particularly beneficial in reducing the chance of 
failure in model fitting. 





## Example 0: Synthetic data

```{r}
set.seed(123)
nn <- 1000
pp <- 50

n_effect_vars <- 3
effect_idx <- sample.int(pp, size = n_effect_vars)
# coefs <- rnorm(n_effect_vars)
coefs <- rep(0.5, times = n_effect_vars)
print(data.frame(effect_vars = effect_idx, effect_size = coefs))

X <- matrix(rnorm(nn*pp), nrow = nn)

Eta <- X[, effect_idx, drop=F] %*% as.matrix(coefs)

Y <- rpois(nn, exp(Eta))
hist(Y)

res_cl <- clamp(X, Y, family = "poisson",
                standardize = F)
summarize_coefficients(res_cl)
```

